Chemical Tagging Using Minimum Message Length. I.\\ Gaussian Mixture Modelling

### 2017-04-20 articles/chemical-tagging-gmm/ms.tex
the model

\usepackage{bm}
\usepackage{verbatim} % for \begin{comment}'ing out sections.

@@

% Common terms
\def\teff{T_{\rm eff}}
\def\logg{\log{g}}

% Have some maaaath.
\def\infinity{\rotatebox{90}{8}}

@@

- Dimensionality of chemical abundance space.

@@

\begin{itemize}
\item We assume that the data (e.g., the $D$ detailed chemical abundances of 
      $N$ stars) can be represented by a mixture of $K$ multivariate Gaussian
      distributions.
\item We assume that there are no missing data values. For example, if a star
      has been observed, then there is a complete set of $D$ chemical 
      abundances available.
\item We will assume that the number of \emph{true} multivariate Gaussian
      distributions $K_{true}$ is not known. This assumption applies to our 
      experiements involving generated data, and those using real data.
\item In addition to not knowing the \emph{number} of multivariate normal
      distributions, we further assume that the means $\bm{\mu}$ and 
      covariance matrices $\bm{C}$ of each $K$th mixture is unknown.
\item We assume that the covariance matrix of individual components can be
      described as `free' or `full', insomuch that the data within a cluster
      can have off-diagonal correlation terms, and those terms are unknown.
\item In all experiments we will assume that the data have homoskedastic noise
      properties.
\item The relative weights $\bm{w}$ (or mixing proportions) of $K$ 
      distributions is unknown. 
\item We assume a relative (or conditional) probability distribution for each 
      star belonging to a given cluster. That is to say that we do not adopt a 
      `\emph{hard selection}' approach to cluster modelling, where stars would
      be assigned as definitively belonging to one cluster or another. Our
      approach could be described as `\emph{soft} clustering'. This assumption
      has practical numerical outcomes and implications for the message length.
\end{itemize}

We have made relatively few explicit assumptions about the model for the data,
but even the model described constitutes a very formidable model selection 
problem. This is because we do not know \emph{how many} multivariate Gaussian
distributions we ought to chose to fit the data. Including more mixtures will
consistently provide a better fit to the data (in a likelihood sense), but we 
must make some decision or heuristic as to whether the additional mixture is 
justified.

The probability density function $f$ for a multivariate normal distribution 
with $d$ dimensions is given by

\begin{equation}
    f(\bm{y}|\bm{\mu},\bm{\Sigma}) 
        = \frac{1}{\sqrt{(2\pi)^d|\bm{\Sigma}|}}
          \exp{\left[-\frac{1}{2}(\bm{y} - \bm{\mu})^\intercal\bm{\Sigma}^{-1}(\bm{y} - \bm{\mu})\right]}
\end{equation}

\noindent{}where $\bm{\mu}$ and $\bm{\Sigma}$ is the multivariate mean and 
covariance matrix, and $\bm{y}$ are the data.

% Full probability for the data.

When modelling a mixture of
Gaussians, there are two approaches in determining assignments (or
responsibilities) of each star to a given cluster. One approach is to include
a (discrete) model parameter for every data point, which represents the index
of the mixture that the observation is assigned to. The second approach is to 
use a weighted mixture. These two are mathematically identical.
% TODO

\begin{equation}
% TODO
\end{equation}

>>>>

\usepackage{bm}
%\usepackage{boldsymbol}
\usepackage{verbatim} % for \begin{comment}'ing out sections.

@@

% Common terms
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}

\def\teff{T_{\rm eff}}
\def\logg{\log{g}}

% Have some maaaath.
\def\infinity{\rotatebox{90}{8}}
\def\veccov{\vect{C}}
\def\vecmean{\vect{\mu}}
\def\vectheta{\vect{\theta}}
\def\weight{w}
\def\weights{\vect{\weight}}
\def\datum{y}
\def\data{\vect{\datum}}

@@

% TODO
- Dimensionality of chemical abundance space.

@@

\begin{itemize}
  \item We assume that the data (e.g., $D$ detailed chemical abundances for
        $N$ stars) can be represented by a mixture of $K$ multivariate 
        Gaussian distributions.
  \item We assume that there are no missing data values. 
        For example, if a star has been observed, then there is a complete set
        of $D$ chemical abundances available.
  \item We will assume that the number of \emph{true} multivariate Gaussian
        distributions $K_{true}$ is not known. 
        This assumption applies to our experiements involving generated data, 
        and those using real data.
  \item In addition to not knowing the \emph{number} of multivariate normal
        distributions, we further assume that the means $\vecmean$ and 
        covariance matrices $\veccov$ of each $K$-th mixture is unknown.
  \item We assume that the covariance matrix of individual components can be
        described as `free' or `full', insomuch that the data within a cluster
        can have off-diagonal correlation terms, and those terms are unknown.
  \item In all experiments we will assume that the data have homoskedastic 
        noise properties.
  \item The relative weights $\weights$ (or mixing proportions) of the $K$ 
        distributions is unknown. 
  \item We assume a relative (or conditional) probability distribution for 
        each star belonging to a given cluster. That is to say that we do not 
        adopt a `\emph{hard selection}' approach to cluster modelling, where 
        stars would be assigned as definitively belonging to one cluster or 
        another. 
        Our approach could be described as `\emph{soft} clustering'.
        This assumption has practical implications for optimization, and for
        the total message length.
\end{itemize}

\noindent{}The probability density function $p$ for a multivariate normal 
distribution with $D$ dimensions is given by

\begin{equation}
  p(\data|\vecmean,\veccov) 
      = \frac{1}{\sqrt{(2\pi)^d|\veccov|}}
        \exp{\left[-\frac{1}{2}(\data - \vecmean)^\intercal\veccov^{-1}(\data - \vecmean)\right]}
\end{equation}

\noindent{}where $\vecmean$ and $\veccov$ is the multivariate mean and 
covariance matrix, and $\data$ are the data.
For a fixed number of $K$ multivariate Gaussian mixtures, the probability
density function can be written as

\begin{equation}
  p(\data|\vectheta) = \sum_{k=1}^{K} \weight_k \, p(\data|\vectheta_k)
\end{equation}

\noindent{}where $\vectheta_k \equiv \{\vecmean_k, \veccov_k\}$,
$\vectheta \equiv \{\vectheta_1,\dots,\vectheta_K,\weight_1,\dots,\weight_K\}$, and
$\weight_k$ is the relative weight (or the mixing probability) of the $k$-th 
mixture,\footnote{
  An alternative representation would be to include integer variables for each
  of the $N$ stars, where each variable represents the cluster which that star
  belongs.
  These two models are mathematically identical \citep{Foreman_Mackey_2014},
  but the integer representation is highly dimensional and difficult to
  optimize.}
and we require that $\sum_{k=1}^{K} \weight_k = 1$.
Although the true number of $K$ mixtures is unknown (as are the means $\vecmean$
and covariance matrices $\veccov$), we will address the selection of $K$
and parameter estimates of $\vectheta$ in Section \ref{sec:fitting-the-data}.
